\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-05 analysis/paper/paper.Rmd
add analysis of stanford data

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill Univeristy
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: http://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill Univeristy
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

# Code formatting {.unnumbered}

Don't use markdown, instead use the more precise latex commands:

* \proglang{Java}
* \pkg{plyr}
* \code{print("abc")}

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- FALSE
```

@@

As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogenous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling. This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base samping *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

@@

The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling. We explicitely aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our unit tests.

@@

As espected, we see that splines model provides a better fit.

@@

>>>>

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: http://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 7
fig_height: 6
fig_caption: true
---

<!-- # Code formatting {.unnumbered} -->

<!-- Don't use markdown, instead use the more precise latex commands: -->

<!-- * \proglang{Java} -->
<!-- * \pkg{plyr} -->
<!-- * \code{print("abc")} -->

@@

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- TRUE
```

@@

As shown by Saarela & Arjas [-@saarela2015non] (and further expanded in Saarela [-@saarela2016case]), writing the likelihood arising from this data-generating mechanism using the framework of non-homogeneous Poisson processes, we eventually reach an expression where each person-moment's contribution is of the form
$$\frac{h(t)^{dN(t)}}{\rho(t) + h(t)},$$
where $N(t)$ is the counting process associated with the event of interest, $h(t)$ is the corresponding hazard function, and $\rho(t)$ is the hazard function for the Poisson process associated with case-base sampling. This parametric form suggests that we can readily estimate log-hazards of the form $\log(h(t)) = g(t; X)$ using logistic regression, where each observation corresponds to a person moment, the function $g(t; X)$ is linear in a finite number of parameters, and where we treat $-\log(\rho(t))$ as an offset.

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

@@

The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling. We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our unit tests.

@@

As expected, we see that splines model provides a better fit.

@@

# Case study 4--Stanford Heart Transplant Data

Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share these properties. To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.

```{r stanford-poptime, eval = eval_cs4}
library(survival)
library(casebase)

stanford_popTime <- popTime(jasa, time = "futime", 
                            event = "fustat")
plot(stanford_popTime)
```

Since the exposure is time dependent, we need to manually define the exposure variable *after* case-base sampling and *before* fitting the hazard function. For this reason, we will use the `sampleCaseBase` function directly.

```{r, eval = eval_cs4}
library(tidyverse)
library(lubridate)

cb_data <- sampleCaseBase(jasa, time = "futime", 
                          event = "fustat", ratio = 10)
```

Next, we will compute the number of days from acceptance into the program to transplant, and we use this variable to determine whether each population-moment is exposed or not.

```{r, eval = eval_cs4}
# Define exposure variable
cb_data <- mutate(cb_data,
                  txtime = time_length(accept.dt %--% tx.date, 
                                       unit = "days"),
                  exposure = case_when(
                    is.na(txtime) ~ 0L,
                    txtime > futime ~ 0L,
                    txtime <= futime ~ 1L
                  ))
```

Finally, we can fit the hazard using various linear predictors.

```{r, eval = eval_cs4, warning = FALSE}
library(splines)
# Fit several models
fit1 <- fitSmoothHazard(fustat ~ exposure,
                        data = cb_data, time = "futime")
fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
                        data = cb_data, time = "futime")
fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
                        data = cb_data, time = "futime")
fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
                        data = cb_data, time = "futime")
```

Note that the fourth model (i.e. `fit4`) includes an interaction term between exposure and follow-up time. In other words, this model no longer exhibit proportional hazards. We can then compare the goodness of fit of these four models using the Akaike Information Criterion (AIC).

```{r, eval = eval_cs4}
# Compute AIC
c("Model1" = AIC(fit1),
  "Model2" = AIC(fit2),
  "Model3" = AIC(fit3),
  "Model4" = AIC(fit4))
```

As we can, the best fit is the fourth model. By visualizing the hazard functions for both exposed and unexposed individuals, we can more clearly see how the hazards are no longer proportional.

```{r stanford-hazard, eval = eval_cs4}
# Compute hazards---
# First, create a list of time points for both exposure status
hazard_data <- expand.grid(exposure = c(0, 1),
                           futime = seq(0, 1000,
                                        length.out = 100))
# Set the offset to zero
hazard_data$offset <- 0 
# Use predict to get the fitted values, and exponentiate to 
# transform to the right scale
hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
                                 type = "link"))
# Add labels for plots
hazard_data$Status = factor(hazard_data$exposure,
                            labels = c("NoTrans", "Trans"))

ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
    geom_line() +
    theme_minimal() +
    theme(legend.position = 'top') +
    ylab('Hazard') + xlab('Follow-up time')
```

The non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.

```{r stanford-risk, eval = eval_cs4}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit4, newdata = newdata, 
                        time = seq(0, 1000, length.out = 100))

colnames(absrisk) <- c("Time", "NoTrans", "Trans")

# Rearrange the data
absrisk <- gather(as.data.frame(absrisk),
                  "Status", "Risk", -Time)

ggplot(absrisk, aes(Time, Risk, colour = Status)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = 'top') +
  expand_limits(y = 0.5) +
  xlab('Follow-up time') + ylab('Cum. Incidence')
```

Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 30 days.

```{r stanford-1y-haz, eval = eval_cs4}
# Compute hazards---
# First, create a list of time points for both exposure status
one_yr_haz <- data.frame(futime = seq(0, 365,
                                      length.out = 100))
one_yr_haz <- mutate(one_yr_haz,
                     offset = 0,
                     exposure = if_else(futime < 30, 0, 1))
one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
                                type = "link"))
ggplot(one_yr_haz, aes(futime, hazard)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = 'top') +
  ylab('Hazard') + xlab('Follow-up time') +
  geom_vline(xintercept = 30, linetype = 'dashed')
```

We can then compare the 1-year mortality risk without transplant and with transplant at 30 days.

```{r stanford-1y-risk, eval = eval_cs4}
absoluteRisk(fit4, newdata = data.frame(exposure = 0),
             time = 365)

# Use the trapezoidal rule to estimate the cumulative hazard
min_hazard <- min(one_yr_haz$hazard)
max_hazard <- max(one_yr_haz$hazard)
increment <- 365/99
cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
1 - exp(-cumhaz_est)
```

As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 30 days.
